---
title: "SMPA - Grouping Assignment"
author: "Avril Silva"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r}

#load libraries
library(tidyverse)
library(lubridate)

#load data
payments <- readRDS("data/trump_campaign_payments.rds")

```

## Introduction

**What data will you be working with?**

A dataset of disbursements (payments and purchases) by the Donald J. Trump for President campaign committee during the 2020 election cycle. The included payments are from 2017 and 2018.

This data and examples were developed as part of the NYT Open training, which we've adapted from here. Note that the dataset has been alterted for instructional purposes, so doesn't represent the original FEC data in its entirety.

In the environment tab in the upper right, click on the `payments` dataset to open the RStudio viewer to examine the data. You can also run the chunk below as well to bring up a slice of the data.

```{r}

head(payments)


```

## Before you begin

Take a moment to look at the various categories included for each disbursement.

-   Notice that there are several date-related fields:

    -   FEC election cycle

    -   report year

    -   election type

    -   disbursement date

-   There are a few fields about the recipient. Along with city and state, you have:

    -   the recipient name

    -   the payee's first and last name (an individual who was paid for the "recipient") and

    -   the entity type (individual v. organization)

## Initial Narrative Questions

Without doing any analysis yet, what are three questions you might try to answer using this dataset? Write your narrative answers in the spaces below.

1.  What organization had the most disbursements paid to in 2017? Top three?  

2.  How many different payees are listed in 2018?

3.  What are the top three individuals that were paid disbursements in 2018?

## Coding Assignment Questions

**Please write your R tidyverse code to help answer the questions below in the provided R chunks. You can make more than one chunk if you prefer.**

**Also, describe what you're doing either in text out "in the open" above the chunk, or in comments (prefaced by #) in the R chunk itself.**

**You'll need to use grouping, filtering, or a combination of both to find the answers. Remember that R is case-sensitive when you refer to any text values within the columns**

<br>

How much was spent in the 2017 vs. 2018 reporting year?

```{r}

payments %>% 
  group_by(report_year) %>% 
  summarise(tot_spent = sum(disbursement_amount))

# To differentiate between how much was spent between 2017 and 2018, I made sure to group the values by the reported year and then make a new field for the total spent in each year by using the sum function to add up all the disbursements. Using this, you can find that the spending was more in 2018. 



```

Tally the total overall disbursement amounts for each month (distinguishing between say, Oct. 2018 v. Oct. 2019). In which month was spending the highest?

```{r}

payments %>% 
  group_by(disbursement_month, disbursement_year) %>% 
  summarise(tot_spent = sum(disbursement_amount)) %>% 
  arrange(desc(tot_spent))

# In order to organize it by each month of each year in a way that shows me the values from largest to smallest, I broke it up bit by bit, so I started with the grouping command to make sure I had the year and month grouped up and then I did the same summarise command as the previous question and then sorted it using the arrange command to do it in descending order. As a result, you can see that the most spending occurred in October of 2018.                                                             


```

Which recipient (recipient_name) was paid the most in the 2020 election cycle?

```{r}

payments %>% 
  group_by(recipient_name) %>% 
  summarise(tot_spent = sum(disbursement_amount)) %>% 
  arrange(desc(tot_spent))

# Using the same grouping, summarise, and arrange tools as the previous questions, I used the same tot_spent sum field to add up all the disbursements of each recipient and do the same sorting to show the results in a descending order. This one was a bit easier than the last one because of the more practice I was able to have. As you can see, the recipient who was paid the most in the 2020 election cycle, was "AMERICAN MADE MEDIA CONSULTANTS, LLC"



```

Look at what this recipient was paid for. What service or good (see the "disbursement_description") were they paid the most for?

```{r}

payments %>% 
  filter(recipient_name == "AMERICAN MADE MEDIA CONSULTANTS, LLC") %>% 
  group_by(disbursement_description) %>% 
  summarise(tot_spent = sum(disbursement_amount)) %>% 
  arrange(desc(tot_spent))

# Now that I have gotten a little bit of a better hang of the commands, knowing to filter for the specific organization was all I needed to know and then the rest came into place after the disbursement description field was recognized. This code gets easier knowing I can build off of what I did before. AMERICAN MADE MEDIA CONSULTANTS, LLC, as seen in the data, was paid the most for what the data recognizes as PLACED MEDIA. 


```

What's the most common category overall (see "disbursement_description") the campaign spent money on?

```{r}

payments %>% 
  group_by(disbursement_description) %>% 
  summarise(tot_spent = sum(disbursement_amount)) %>% 
  arrange(desc(tot_spent))


# This question was even easier than the last one because I didn't even need the filter and could just group by the disbursement description to find that the most common category overall the campaign spent money on was what they claim to be "DIGITAL CONSULTING/ONLINE ADVERTISING". 


```

Who were the 5 biggest recipients of campaign spending on "LEGAL CONSULTING"..?

```{r}

payments %>% 
  filter(disbursement_description == "LEGAL CONSULTING") %>% 
  group_by(recipient_name) %>% 
  summarise(tot_spent = sum(disbursement_amount)) %>% 
  arrange(desc(tot_spent)) %>% 
  head(5)

# Filtering for the LEGAL CONSULTING category under the disbursement description field, I used the same grouping by recipient name I did a few questions ago and the same summarizing and arranging lines that I have used throughout the problem set and then just selected the top five using the head command. Using these commands, I was able to find that the five biggest recipients of campaign spending under LEGAL CONSULTING were JONES DAY, LAROCCA HORNIK ROSEN GREENBERG & BLAHA LLP, KASOWITZ, BENSON, TORRESS & FRIEDMAN, THE LAW OFFICES OF ALAN S. FUTERFAS, and MCDERMOTT WILL & EMERY.  




```


Who are the top 5 individuals (vs. organization) were paid by the campaign committee?

```{r}

payments %>% 
  filter(entity_type == "IND") %>% 
  group_by(recipient_name) %>% 
  summarise(tot_spent = sum(disbursement_amount)) %>% 
  arrange(desc(tot_spent)) %>% 
  head(5)

payments %>% 
  filter(entity_type == "ORG") %>% 
  group_by(recipient_name) %>% 
  summarise(tot_spent = sum(disbursement_amount)) %>% 
  arrange(desc(tot_spent)) %>% 
  head(5)

# To find the top five individual recipients and the top five organizations, I used essentially the same code for both, using the same filtering, grouping, summarizing, and arrangement lines I used earlier. As a result, I was able to find that the top five individual recipients were Michael Glassner, Sean Dollman, Darren Centinello, and Christopher Carr and John Pence, and the top five organizations were AMERICAN MADE MEDIA CONSULTANTS, LLC, PARSCALE STRATEGY, LLC, JONES DAY, AMERICAN EXPRESS, and GILES-PARSCALE. 



```


For the indvidual who was paid the most, which payment category (disbursement_description) was the top one for him or her? And how much money were they paid in that category?

```{r}

payments %>% 
  filter(recipient_name == "GLASSNER, MICHAEL") %>% 
  group_by(disbursement_description) %>% 
  summarise(tot_spent = sum(disbursement_amount)) %>% 
  arrange(desc(tot_spent))

# Considering we only had to filter for the one individual, I was able to use essentially the lines from before, this time just grouping the disbursement descriptions and finding their sums to find that the top individual, Michael Glassner, was paid 499000.00 for his payroll, which was his top payment category. 



```
